Formal Analysis of Rate-Monotonic Scheduling Implementation via Real-Time Maude

### 2015-09-26 RMS/submission.tex
finish section 4.5

\begin{algorithm}
  \caption{The C-Like Pseudocode of $schedule()$}
  \label{a:schedule}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts}
  \State \Call{$updateStatus$}{$taskList$};
  \State $timer = timer + 1$; \label{l:timer}
  \State $p = taskList$;
  \While{$p$} 
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return;
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts}
      \State $p\rightarrow function()$; \Comment{to run the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
	\State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\end{algorithm}

@@

We also need a sort \verb|Interval| denoting the interval, within
which the computation time of a task varies, and a sort \verb|Counter|
to represent how long some task has run and how long it should run.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

@@

\subsection{Interrupt Handling}
\label{ss:inthandling}
When an interrupt request comes, it may not be ``seen'' immediately by
the system. It requires the \verb|mask| flag to be cleared. Once the
request is seen, it is handled by performing two stages: the interrupt
mechanism of the hardware (such as clearing \verb|ir|, pushing context
into stack and so on) and running the $schedule()$ function. This is
modeled by the following instanous rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The operation \verb|_.interrupt| models
the interrupt mechanism performed by the hardware and does four
things: (i)clearing the \verb|ir| flag, which means the request has
been handled; (ii)pushing the current \verb|pc| value into the stack,
storing the interrupted context; (iii)assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that $schedule()$ is
running; and (iv)setting the \verb|mask| flag, to mask the coming
interrupt requests.

It is worth pointing out that \emph{scheduling} in our model does not
refer to the stage in which $schedule()$ is running. We use it to
refer to the stage scheduling the first should-be-run periodic task,
i.e.  the code strictly before Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore,
\begin{definition}
  The \emph{scheduling time} talked about in this paper consists of
  the time for switching from a running task to the function
  $schedule()$ when handling interrupts, the time spent by
  $schedule()$ finding and setting the \emph{first} should-be-run
  periodic task, and the time for switching from $schedule()$ to that
  first should-be-run task.
\end{definition}
Similarly, \emph{switching} in our model refers to the stage finding
out the next should-be-run periodic task when one is finished.
\begin{definition}
  The \emph{switching time} in this paper consists of the time for
  switching from a finished task back into $schedule()$, the time
  spent by $schedule()$ finding and setting the \emph{next}
  should-be-run periodic task, and the time for switching from
  $schedule()$ to that next should-be-run task.
\end{definition}

>>>>

\begin{algorithm}
  \caption{The C-Like Pseudocode of $schedule()$}
  \label{a:schedule}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts}
  \State \Call{$updateStatus$}{$taskList$}; \label{l:updatestatus}
  \State $timer = timer + 1$; \label{l:timer} \label{l:inc}
  \State $p = taskList$;
  \While{$p$} \label{l:startrun1st}
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return;
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts} \label{l:endrun1st}
      \State $p\rightarrow function()$; \Comment{to run the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
	\State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\end{algorithm}

@@

We also need a sort \verb|Interval| denoting the interval, within
which the computation time of a task varies and which we call
\emph{possible finishing interval}, and a sort \verb|Counter| to
represent the time some task has been running and the corresponding
possible finishing interval.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request comes, it may not be ``seen'' immediately by
the system. It requires the \verb|mask| flag to be cleared. Once the
request is seen, it is handled by performing two stages: the interrupt
mechanism of the hardware (such as clearing \verb|ir|, pushing context
into stack and so on) and running the $schedule()$ function. This is
modeled by the following instanous rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The operation \verb|_.interrupt| models
the interrupt mechanism performed by the hardware and does four
things: (i)clearing the \verb|ir| flag, which means the request has
been handled; (ii)pushing the current \verb|pc| value into the stack,
storing the interrupted context; (iii)assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that $schedule()$ is
running; and (iv)setting the \verb|mask| flag, to mask the coming
interrupt requests. As for \verb|_.startScheduling|, it performs
updating the status of $taskList$ and increasing $timer$ by $1$,
corresponding to Line~\ref{l:updatestatus} and \ref{l:inc} in
Algorithm~\ref{a:schedule}.
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
       = ((updateStatus L with T) inc(T) STS HW ISRC) .
\end{verbatim}

It is worth pointing out here that \emph{scheduling} in our model does
not refer to the stage in which $schedule()$ is running. We use it to
refer to, for simplicity, the stage scheduling the first should-be-run
periodic task, i.e.  the code strictly before Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore,
\begin{definition}
  The \emph{scheduling time} talked about in this paper consists of
  the time for switching from a running task to the function
  $schedule()$ when handling interrupts, the time spent by
  $schedule()$ finding and setting the \emph{first} should-be-run
  periodic task, and the time for switching from $schedule()$ to that
  first should-be-run task.
\end{definition}
Similarly, \emph{switching} in our model refers to the stage finding
out the next should-be-run periodic task when one is finished.
\begin{definition}
  The \emph{switching time} in this paper consists of the time for
  switching from a finished task back into $schedule()$, the time
  spent by $schedule()$ finding and setting the \emph{next}
  should-be-run periodic task, and the time for switching from
  $schedule()$ to that next should-be-run task.
\end{definition}

With our terminology, when the running time of the system task
\verb|scheduling| reaches its possible finishing interval, the
following rule may apply to finish the scheduling stage:
\begin{verbatim}
  crl [scheduling-finish] :
    (L T STS HW ISRC) => (SYSTEM).finishScheduling
    if SYSTEM := (L T STS HW ISRC) 
       /\ (SYSTEM).running == scheduling 
       /\ scheduling mayFinish?in STS .
\end{verbatim}
where operation \verb|_.running| returns the current \verb|pc| value
of the system and \verb|_mayFinish?in_| checks if the running time of
the task reaches its possible finishing interval:
\begin{verbatim}
  op _mayFinish?in_ : Oid SysTasks ~> Bool .
  eq O mayFinish?in [ < O : SysTask | cnt : C > REST ] 
       = C mayFinish? .
  op _mayFinish? : Counter -> Bool .
  eq [ R / [ MIN , MAX ] ] mayFinish?
       = if R lt MIN then false else true fi .
\end{verbatim}
And the operation \verb|_.finishScheduling| performs the remaining
instructions, which are left by \verb|_.startScheduling|, in the
scheduling stage: finding the first should-be-run periodic task and
setting it to run (Line~\ref{l:startrun1st}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}).
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
       = (L T (finish scheduling in STS) HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Line~\ref{l:startrun1st}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or letting it run respectively.

Similar to \verb|scheduling|, the \verb|switching| stage starts when
the running time of the current running periodic task reaches its
possible finishing interval and finishes when its own running time
does so, with the two following similar rules:
\begin{verbatim}
  crl [task-finish] :
    (L T STS HW ISRC) => (SYSTEM).startSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ some N := (SYSTEM).running
       /\ some N mayFinish?in L .
  crl [switching-finish] :
    (L T STS HW ISRC) => (SYSTEM).finishSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ (SYSTEM).running == switching
       /\ switching mayFinish?in STS .
\end{verbatim}
